---
title: "autos Backup"
author: "Zach"
date: 2018-09-04
output:
  github_document:
  toc: true
---
  
The following is a pre-baked analysis of the UCI `autos` dataset.

```{r setup, echo=FALSE, warning=FALSE}
library(tidyverse)
library(tidyPca)
library(ggrepel)

## Data filename 
file_name <- "~/Sync/data/autos/imports-85.data"
## Variable names from the UCI site
var_names <- 
  c(
    "symboling",          # Actuarial risk adjustment: -3 Safest, +3 Riskiest
    "normalized_losses",  # Average insurance loss, averaged over body_style     
    "make",               # Manufacturer
    "fuel_type",          # Gas or diesel 
    "aspiration",         # Standard or turbo      
    "num_of_doors",       #      
    "body_style",         # hardtop, wagon, sedan, hatchback, convertible      
    "drive_wheels",       # Front, rear, all      
    "engine_location",    # Front, rear      
    "wheel_base",         # Distance between front & rear wheels      
    "length",             #      
    "width",              #      
    "height",             #      
    "curb_weight",        # Weight w/o cargo      
    "engine_type",        # Engine layout      
    "num_of_cylinders",   #      
    "engine_size",        # Total engine displacement      
    "fuel_system",        # Fuel feed system      
    "bore",               # Cylinder diameter      
    "stroke",             # Cylinder displacement distance      
    "compression_ratio",  # Swept volume compression ratio      
    "horsepower",         # Engine power      
    "peak_rpm",           # Peak rotations per minute (redline?)
    "city_mpg",           # City fuel efficiency      
    "highway_mpg",        # Highway fuel efficiency      
    "price"               # MSRP    
  )

## Helper functions
## --------------------------------------------------
## Do a two-PC plot, colored by an additional variable
do_pca_plot <- function(df, col_var, components = TRUE, scale = 2.0) {
    col_var <- enquo(col_var)

    df_components <-
        df %>%
        pull(pca) %>%
        .[[1]] %>%
        .$rotation %>%
       as_tibble(rownames = "val")

    ## Start the plot
    fig <-
        df %>%
        unnest(data_aug) %>%
        rename_all(~ str_replace(., patter = "\\.fitted", replacement = "")) %>%
        ggplot()

    ## Optional color
    if (rlang::quo_is_missing(col_var)) {
        fig <-
            fig +
            geom_point(aes(PC1, PC2))

    } else {
        fig <-
            fig +
            geom_point(aes(PC1, PC2, color = as.factor(!! col_var)))

    }

    ## Optional component plot
    if (components) {
        fig <-
            fig +
            geom_segment(
                data = df_components,
                aes(x = 0, y = 0, xend = PC1 * scale, yend = PC2 * scale),
                ## arrow = arrow(),
                color = "red"
            )+
            geom_text_repel(
                data = df_components,
                aes(PC1 * scale, PC2 * scale, label = val),
                color = "red",
                box.padding = 1.0,
                segment.size = 0.1
            )
    }

    ## Finish and return the plot
    fig +
        theme(legend.position = "bottom") +
        labs(
            x = "Principal Component Score 1",
            y = "Principal Component Score 2"
        )
}

parse_numchar <- function(x) {
  case_when(
        x == "two" ~ 2L,
        x == "three" ~ 3L,
        x == "four" ~ 4L,
        x == "five" ~ 5L,
        x == "six" ~ 6L,
        x == "seven" ~ 7L,
        x == "eight" ~ 8L,
        x == "nine" ~ 9L,
        x == "ten" ~ 10L,
        x == "eleven" ~ 11L,
        x == "twelve" ~ 12L,
        TRUE ~ NA_integer_
      )
}
```

## Wrangling
<!-- ------------------------------------------------- -->

The data are essentially in CSV format, though we'll need to do some wrangling.

```{r load_n_wrangle}
df_raw <- 
  read_csv(
    file_name, 
    col_names = var_names,
    col_types = cols(
      .default = col_number(),
      symboling = col_integer(),
      normalized_losses = col_integer(),
      make = col_character(),
      fuel_type = col_character(),
      aspiration = col_character(),
      num_of_doors = col_character(), # Needs manual fix!
      body_style = col_character(),
      drive_wheels = col_character(),
      engine_location = col_character(),
      engine_type = col_character(),
      num_of_cylinders = col_character(), # Also needs manual fix!
      engine_size = col_integer(),
      fuel_system = col_character()
    )
  )
```

The string-coded numbers are obnoxious; let's map those to integers.

```{r}
df_raw %>% pull(num_of_doors) %>% unique()
df_raw %>% pull(num_of_cylinders) %>% unique()
```


```{r}
df_autos <- 
  df_raw %>% 
  mutate(
    num_of_doors = parse_numchar(num_of_doors),
    num_of_cylinders = parse_numchar(num_of_cylinders)
  )
```


## Analysis
<!-- ------------------------------------------------- -->

First, a glimpse.

```{r}
df_autos %>% glimpse()
```

The numeric variables seem to break down into

| Geometry   | `wheel_base`, `length`, `width`, `height`, `curb_weight` |
| Engine     | `engine_size`, `num_of_cylinders`, `engine_size`, `bore`, `stroke`, `compression_ratio`, `horsepower`, `peak_rpm` |
| Efficiency | `city_mpg`, `highway_mpg` |
| Monetary   | `price` |

__q1__ What is the relationship between the geometric variables `length`, `width`, `height`, and `wheel_base`?

```{r}
pca_geom <- 
  df_autos %>% 
  mutate_at(vars(length, width, height, wheel_base), .funs = log10) %>% 
  tidy_pca(length, width, height, wheel_base)

pca_geom %>% 
  unnest(pc_frac)

pca_geom %>% 
  unnest(pc_weights)
```

There's a fair bit of structure here; the leading two components capture about `78`% of the variability.
The leading principal component has positive weight on all variables -- it corresponds to generic 'size'
of a vehicle. PC2 is an approximate aspect ratio* -- it describes increases in height with decreases in width,
and smaller decreases in length. PC3 is another form of aspect ratio

*This isn't a proper aspect ratio, as it is not a true dimensionless quantity. It has units of length to
the `r pca_geom %>% unnest(pc_weights) %>% pull(PC2) %>% sum(.)` power.

```{r}
pca_geom %>% 
  do_pca_plot(body_style)
```

We can see this if we select an interior slice of PC1 (to control 
overall size) and plot height against width.

```{r}
pca_geom %>% 
  unnest(data_aug) %>% 
  mutate_at(vars(width, height), ~10 ^ .) %>% 
  filter(abs(PC1) < 1.5) %>%
  ggplot(aes(width, height, color = body_style)) +
  geom_point() +
  geom_point(
    data = . %>%
      group_by(body_style) %>%
      summarize(
        width = median(width, na.rm = TRUE),
        height = median(height, na.rm = TRUE)
      ),
    size = 5,
    alpha = 1 / 2
  )
```


__q2__ What is the relationship between the engine variables `engine_size`, `bore`, `stroke`, `compression_ratio`, and `num_of_cylinders`?

```{r}
pca_engine <- 
  df_autos %>% 
  filter_at(vars(engine_size, bore, stroke, compression_ratio, num_of_cylinders), all_vars(!is.na(.))) %>% 
  mutate_at(vars(engine_size, bore, stroke, compression_ratio), log10) %>%
  tidy_pca(engine_size, bore, stroke, compression_ratio, num_of_cylinders)

pca_engine %>% 
  unnest(pc_frac)
```

The engine variables are not particularly well-described by the first two PCA compoments. At three we get a decent
reduction.

```{r}
pca_engine %>% 
  unnest(pc_weights) %>% 
  mutate(PC1 = - PC1)
```

Engine size should be directly related to the bore (cylinder diameter), stroke (cylinder swept height), and 
number of cylinders. Let's check this.

```{r}
df_engine <- 
  df_autos %>% 
  mutate(
    cyl_vol = stroke * (bore / 2) ^ 2 * pi,
    engine_vol = cyl_vol * num_of_cylinders,
    resid = engine_size - engine_vol
  )

## Computed fit
df_engine %>% 
  ggplot(aes(engine_size, engine_vol)) +
  geom_point()

## Residual plot
df_engine %>% 
  filter(abs(resid) > 5) %>% 
  ggplot(aes(fct_reorder(make, resid), resid)) +
  geom_jitter(width = 0.2, height = 0) +
  theme(
    axis.text.x = element_text(angle = 315)
  ) +
  labs(
    x = "Make",
    y = "Engine Size Residual"
  )
```

The calculated engine volume is almost perfect; however, there are some outliers. I've not managed to determine
where this discrepancy comes from.

__q3__ What about dimensionality made these questions hard?

1. Choosing subsets of variables required domain-specific knowledge
2. There are many variables to interpret simultaneously.

__q4__ What else is going on in these data?

There's a very funny relationship between vehicle price and efficiency:

```{r}
df_autos %>% 
  ggplot(aes(price)) +
  geom_smooth(aes(y = highway_mpg, color = "Highway")) +
  geom_smooth(aes(y = city_mpg, color = "City")) +
  geom_point(aes(y = highway_mpg, color = "Highway")) +
  geom_point(aes(y = city_mpg, color = "City")) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "Price",
    y = "Efficiency"
  )
```

More expensive cars are less efficient! You get what you pay for, I suppose....
