---
title: "Aircraft Attributes"
author: "ZDR"
date: 2018-08-13
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
---

The following is a glimpse at the FAA Aircraft Characteristic
[https://www.faa.gov/airports/engineering/aircraft_char_database/](database).
These data are used for sizing airport, so the database is likely representative
of commercial aircraft in use today.

```{r setup}
library(tidyverse)
library(readxl)
library(GGally)
library(ggrepel)

## File location
filename_faa <- "~/Documents/data/faa/FAA-Aircraft-Char-Database-v2-201801.xlsx"

## Helper functions
normsum <- function(v) {
    cumsum(v) / sum(v)
}

tidy_pca <- function(data, ..., doCenter = TRUE, doScale = TRUE) {
    require(tidyverse)
    require(broom)

    group_var <- quos(...)

    data %>%
        nest() %>%
        mutate(
            pca = map(data, ~ prcomp(
                               .x %>% select(!!! group_var),
                               center = doCenter,
                               scale = doScale
                           )),
            data_aug = map2(
                pca,
                data,
                ~ augment(.x, data = .y) %>%
                  select(-.rownames) %>%
                    rename_all(~str_replace(., "\\.fitted", ""))
            ),
            pc_weights = map(
                pca,
                ~ .$rotation %>%
                    as_tibble(rownames = "var")
            ),
            pc_frac = map(
                pca,
                ~ .$sdev %>%
                    tibble(sig = ., sig_frac = normsum(.))
            )
        ) %>%
        select(-data)
}
```

There are a number of comments at the top of this Excel FAA data; let's skip all
of that.

```{r load}
df_raw <-
    filename_faa %>%
    read_excel(skip = 4)
```

First, some basic facts.

```{r first_look}
names(df_raw)
summary(df_raw)
```

Some of these ought to be numeric, and the variable names are obnoxious. Let's
do some wrangling.

```{r wrangle}
df_faa <-
    df_raw %>%
    rename_all(str_to_lower) %>%
    rename_all(~str_remove(., pattern = "\\(.*\\)")) %>%
    rename_all(~str_remove(., pattern = ",.*$")) %>%
    rename_all(~str_remove(., pattern = "\\s+$")) %>%
    rename_all(~str_replace_all(., pattern = "\\s+", replacement = "_")) %>%
    rename_all(~str_replace_all(., pattern = "#", replacement = "num")) %>%
    rename_all(~str_replace_all(., pattern = "/", replacement = "_")) %>%
    mutate_at(
        .vars = vars(approach_speed, wingspan, length, tail_height, mtow),
        .funs = as.numeric
    )
```

Let's focus on the numeric variables for now. A set of pair plots will give us a
first impression if there's any structure in the data.

```{r num_pairs}
df_faa %>%
    select(num_engines, approach_speed, wingspan, length, tail_height, mtow) %>%
    ggpairs()
```

It seems there's quite a bit of structure! Dimensional Analysis would suggest
that log-transforming the data might straighten things out.

```{r log_pairs}
df_log <-
    df_faa %>%
    mutate_at(
        .vars = vars(approach_speed, wingspan, length, tail_height, mtow),
        .funs = log10
    ) %>%
    rename_at(
        .vars = vars(approach_speed, wingspan, length, tail_height, mtow),
        .funs = ~str_c("log10_", .)
    )

df_log %>%
    select(
        num_engines,
        log10_approach_speed,
        log10_wingspan,
        log10_length,
        log10_tail_height,
        log10_mtow) %>%
    ggpairs()
```

Now let's run a PCA to identify any low-dimensional structure.

```{r log_pca}
df_pca_log <-
    df_log %>%
    select(
        num_engines,
        log10_approach_speed,
        log10_wingspan,
        log10_length,
        log10_tail_height,
        log10_mtow) %>%
    filter_all(all_vars(!is.na(.))) %>%
    tidy_pca(-num_engines)

df_pca_log %>% pull(pc_frac)
```

About `87%` of the variance is explained by two directions; we can plot this.

```{r plot_pca}
scale <- 2.0

df_components <-
    df_pca_log %>%
    pull(pca) %>%
    .[[1]] %>%
    .$rotation %>%
   as_tibble(rownames = "val")

df_pca_log %>%
    unnest(data_aug) %>%
    rename_all(~ str_replace(., patter = "\\.fitted", replacement = "")) %>%
    ggplot(aes(PC1, PC2)) +
    geom_segment(
        data = df_components,
        aes(x = 0, y = 0, xend = PC1 * scale, yend = PC2 * scale),
        arrow = arrow(),
        color = "red"
    ) +
    geom_text_repel(
        data = df_components,
        aes(PC1 * scale, PC2 * scale, label = val),
        color = "red"
    ) +
    geom_point()
```
