---
title: "Aircraft Characteristics"
author: "ZDR"
date: 2018-08-13
output:
  github_document:
    toc: true
urlcolor: blue
---

The following is a glimpse at the FAA Aircraft Characteristics
[database](https://www.faa.gov/airports/engineering/aircraft_char_database/).
These data are used for sizing airports, and were most recently updated January
2018.

```{r setup, echo=FALSE}
library(tidyverse)
library(readxl)
library(GGally)
library(ggrepel)

## Script parameters ---------------------------------------------------
ext <- ".eps"
asp_ratio <- 4 / 4
img_width <- 5
img_height <- img_width * asp_ratio

## File location
filename_faa <- "~/Sync/data/faa/FAA-Aircraft-Char-Database-v2-201801.xlsx"

## Helper functions
normsum <- function(v) {
    cumsum(v) / sum(v)
}

tidy_pca <- function(data, ..., doCenter = TRUE, doScale = TRUE) {
    require(tidyverse)
    require(broom)

    group_var <- quos(...)

    data %>%
        nest() %>%
        mutate(
            pca = map(data, ~ prcomp(
                               .x %>% select(!!! group_var),
                               center = doCenter,
                               scale = doScale
                           )),
            data_aug = map2(
                pca,
                data,
                ~ augment(.x, data = .y) %>%
                  select(-.rownames) %>%
                    rename_all(~str_replace(., "\\.fitted", ""))
            ),
            pc_weights = map(
                pca,
                ~ .$rotation %>%
                    as_tibble(rownames = "var")
            ),
            pc_frac = map(
                pca,
                ~ .$sdev %>%
                    tibble(sig = ., sig_frac = normsum(.))
            )
        ) %>%
        select(-data)
}
## Custom save function
mysave <- function(name) {
    ggsave(
        filename = str_c("../slides/images/", name, ext),
        height = img_height,
        width = img_width,
        units = "in"
    )
}
## Common theme settings
theme_common <- function() {
  theme_minimal() %+replace%
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(4, 4, 4, 4), size = 16),
    axis.title.y = element_text(margin = margin(4, 4, 4, 4), size = 16, angle = 90),

    aspect.ratio = 1 / 1,

    plot.margin   = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),
    panel.spacing = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),

    legend.text = element_text(size = 10),

    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_line(color = "grey90")
  )
}
```

There are a number of comments at the top of this Excel FAA data; let's skip all
of that.

```{r load, message=FALSE}
df_raw <-
    filename_faa %>%
    read_excel(skip = 4)
```

First, some basic facts.

```{r first_look}
names_orig <- names(df_raw)
names_orig
```

Looking at the Excel file, some of these ought to be numeric. Also, the variable
names are obnoxious. Let's do some wrangling.

```{r wrangle, message=FALSE, warning=FALSE}
df_faa <-
    df_raw %>%
    rename_all(str_to_lower) %>%
    rename_all(~str_remove(., pattern = "\\(.*\\)")) %>%
    rename_all(~str_remove(., pattern = ",.*$")) %>%
    rename_all(~str_remove(., pattern = "\\s+$")) %>%
    rename_all(~str_replace_all(., pattern = "\\s+", replacement = "_")) %>%
    rename_all(~str_replace_all(., pattern = "#", replacement = "num")) %>%
    rename_all(~str_replace_all(., pattern = "/", replacement = "_")) %>%
    rename(
        cmg = cockpit_to_main_gear
    ) %>%
    mutate_at(
        .vars = vars(
            approach_speed,
            wingspan,
            length,
            tail_height,
            mtow,
            cmg,
            mgw,
            max_ramp_max_taxi
        ),
        .funs = as.numeric
    ) %>%
    mutate(
        num_engines = as.integer(num_engines)
    )

df_key <-
    tibble(
        var = names(df_faa),
        description = names_orig
    )

df_key %>%
    knitr::kable()
```

## Full scatterplot matrix
<!-- -------------------------------------------------- -->
Let's focus on the numeric variables for now. A set of pair plots will give us a
first impression if there's any structure in the data.

```{r num_pairs, echo=FALSE, warning=FALSE}
df_faa %>%
    select_if(is.numeric) %>%
    ggpairs(progress = FALSE)
```

It seems there's quite a bit of structure! Dimensional Analysis (and John Tukey)
would suggest that log-transforming the data might straighten things out.

```{r log_pairs, echo=FALSE, warning=FALSE}
df_log <-
    df_faa %>%
    mutate_if(
        .predicate = is.double,
        .funs = log10
    ) %>%
    rename_if(
        .predicate = is.double,
        .funs = ~str_c("log10_", .)
    )

df_log %>%
    select_if(is.numeric) %>%
    ggpairs(progress = FALSE)
```

Now let's run a PCA to identify any low-dimensional structure.

```{r log_pca, echo=FALSE, message=FALSE}
do_pca <- function(df, ...) {
    pca_vars <- quos(...)

    df %>%
    filter_at(
        .vars = vars(!!! pca_vars),
        all_vars(!is.na(.))
    ) %>%
    tidy_pca(!!! pca_vars)
}

df_pca_log <-
    df_log %>%
    do_pca(
        log10_approach_speed,
        log10_wingspan,
        log10_length,
        log10_tail_height,
        log10_mtow
    )
```

```{r pca_report}
df_pca_log %>% pull(pc_frac)
```

About `87%` of the variance is explained by two directions; this is something we
can plot.

```{r fcn_plot_pca, echo=FALSE, message=FALSE}
do_pca_plot <- function(df, col_var, scale = 2.0) {
    col_var <- enquo(col_var)

    df_components <-
        df %>%
        pull(pca) %>%
        .[[1]] %>%
        .$rotation %>%
       as_tibble(rownames = "val")

    df %>%
        unnest(data_aug) %>%
        rename_all(~ str_replace(., patter = "\\.fitted", replacement = "")) %>%
        ggplot(aes(PC1, PC2, color = as.factor(!! col_var))) +
        geom_point() +
        geom_segment(
            data = df_components,
            aes(x = 0, y = 0, xend = PC1 * scale, yend = PC2 * scale),
            ## arrow = arrow(),
            color = "red"
        ) +
        geom_text_repel(
            data = df_components,
            aes(PC1 * scale, PC2 * scale, label = val),
            color = "red",
            box.padding = 1.0,
            segment.size = 0.1
        ) +
        theme(legend.position = "bottom")
}
```

```{r pca_plot}
df_pca_log %>%
    do_pca_plot(aac)
```

[Approach
speed](https://www.skybrary.aero/index.php/Approach_Speed_Categorisation) is
legally stratified into classes A-E; a fact that is visible in the figure above.
The first principal component is strongly dependent on the aircraft geometry,
while the second is quite aligned with the approach speed.

```{r approach_vs_engines}
df_faa %>%
    ggplot(aes(num_engines, approach_speed, group = num_engines)) +
    geom_boxplot() +
    geom_text_repel(
        data = df_faa %>% filter(approach_speed > 150, num_engines == 1L),
        mapping = aes(label = model)
    )
```

As we might expect, the approach speed varies with the number of engines. The
fastest (on approach) single-engine aircraft are military.

```{r pca_2}
df_pca_2 <-
    df_log %>%
    do_pca(
        log10_wingspan,
        log10_length,
        log10_cmg,
        log10_mtow,
        log10_tail_height,
        log10_mgw
    )

df_pca_2 %>% pull(pc_frac)

df_pca_2 %>%
    do_pca_plot(num_engines)
```

Many of these interrelations can be derived from scaling laws. For example, lift
arises from a pressure differential, so it tends to scale proportional to wing
area. Area has units of length squared; if we assume a constant aspect ratio
across all aircraft (false, but certainly bounded), then the wing area scales as
the wingspan squared. Maximum takeoff weight (MTOW) scales proportional to lift.
Thus we may expect a roughly quadratic relation between wingspan and MTOW.

```{r lift}
df_log %>%
    ggplot(aes(log10_wingspan, log10_mtow, color = log10_length)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    geom_abline(slope = 2, intercept = 1, linetype = 2)
```

Here we see the power-law fit captures a large part of the trend; coloring the
points by aircraft length suggests that other geometric variations (i.e. varying
aspect ratio) account for part of the misfit.

## Presentation helpers
<!-- -------------------------------------------------- -->

Art recommended presenting individual pair plots; this will be helpful for
pedagogical purposes.

```{r wingspan_v_length}
df_faa %>%
    ggplot(aes(length, wingspan)) +
    geom_point() +
    theme_common()
mysave("faa_wingspan_v_length")
```

```{r wingspan_v_mtow}
df_faa %>%
    ggplot(aes(mtow, wingspan)) +
    geom_point() +
    theme_common()
mysave("faa_wingspan_v_mtow")
```

```{r wingspan_v_approach_speed}
df_faa %>%
    ggplot(aes(approach_speed, wingspan)) +
    geom_point() +
    theme_common()
mysave("faa_wingspan_v_approach_speed")
```

Now let's aggregate these into a scatterplot matrix.

```{r min_scatter}
df_faa %>%
    select(wingspan, length, mtow, approach_speed) %>%
    ggpairs()
mysave("faa_min_scatter")
```

Let's demonstrate how a log transform straightens things out.

```{r min_scatter_log}
df_faa %>%
    select(wingspan, length, mtow, approach_speed) %>%
    mutate_all(log10) %>%
    ggpairs()
mysave("faa_min_scatter_log")
```

Finally, let's do an example PCA.

```{r min_pca}
df_min_pca <-
    df_faa %>%
    filter(
        !is.na(wingspan),
        !is.na(length),
        !is.na(tail_height),
        !is.na(cmg),
        !is.na(mgw),
        !is.na(mtow),
        !is.na(approach_speed),
    ) %>%
    tidy_pca(wingspan, length, mtow, approach_speed)

df_min_pca %>%
    pull(pc_frac) %>%
    knitr::kable()
```

```{r min_pca_plot}
df_min_pca %>%
    do_pca_plot(num_engines)
mysave("faa_min_pca_plot")
```
