---
title: "Aircraft Attributes"
author: "ZDR"
date: 2018-08-13
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
urlcolor: blue
---

The following is a glimpse at the FAA Aircraft Characteristic
[database](https://www.faa.gov/airports/engineering/aircraft_char_database/).
These data are used for sizing airports and was updated January 2018, so the
database is likely representative of commercial aircraft in use today.

```{r setup, echo=FALSE}
library(tidyverse)
library(readxl)
library(GGally)
library(ggrepel)

## File location
filename_faa <- "~/Documents/data/faa/FAA-Aircraft-Char-Database-v2-201801.xlsx"

## Helper functions
normsum <- function(v) {
    cumsum(v) / sum(v)
}

tidy_pca <- function(data, ..., doCenter = TRUE, doScale = TRUE) {
    require(tidyverse)
    require(broom)

    group_var <- quos(...)

    data %>%
        nest() %>%
        mutate(
            pca = map(data, ~ prcomp(
                               .x %>% select(!!! group_var),
                               center = doCenter,
                               scale = doScale
                           )),
            data_aug = map2(
                pca,
                data,
                ~ augment(.x, data = .y) %>%
                  select(-.rownames) %>%
                    rename_all(~str_replace(., "\\.fitted", ""))
            ),
            pc_weights = map(
                pca,
                ~ .$rotation %>%
                    as_tibble(rownames = "var")
            ),
            pc_frac = map(
                pca,
                ~ .$sdev %>%
                    tibble(sig = ., sig_frac = normsum(.))
            )
        ) %>%
        select(-data)
}
```

There are a number of comments at the top of this Excel FAA data; let's skip all
of that.

```{r load, message=FALSE}
df_raw <-
    filename_faa %>%
    read_excel(skip = 4)
```

First, some basic facts.

```{r first_look}
names(df_raw)
```

Looking at the Excel file, some of these ought to be numeric. Also, the variable
names are obnoxious. Let's do some wrangling.

```{r wrangle, message=FALSE, warning=FALSE}
df_faa <-
    df_raw %>%
    rename_all(str_to_lower) %>%
    rename_all(~str_remove(., pattern = "\\(.*\\)")) %>%
    rename_all(~str_remove(., pattern = ",.*$")) %>%
    rename_all(~str_remove(., pattern = "\\s+$")) %>%
    rename_all(~str_replace_all(., pattern = "\\s+", replacement = "_")) %>%
    rename_all(~str_replace_all(., pattern = "#", replacement = "num")) %>%
    rename_all(~str_replace_all(., pattern = "/", replacement = "_")) %>%
    rename(
        cmg = cockpit_to_main_gear
    ) %>%
    mutate_at(
        .vars = vars(
            approach_speed,
            wingspan,
            length,
            tail_height,
            mtow,
            cmg,
            mgw,
            max_ramp_max_taxi
        ),
        .funs = as.numeric
    ) %>%
    mutate(
        num_engines = as.integer(num_engines)
    )
```

Let's focus on the numeric variables for now. A set of pair plots will give us a
first impression if there's any structure in the data.

```{r num_pairs, echo=FALSE, warning=FALSE}
df_faa %>%
    select_if(is.numeric) %>%
    ggpairs(progress = FALSE)
```

It seems there's quite a bit of structure! Dimensional Analysis (and John Tukey)
would suggest that log-transforming the data might straighten things out.

```{r log_pairs, echo=FALSE, warning=FALSE}
df_log <-
    df_faa %>%
    mutate_if(
        .predicate = is.double,
        .funs = log10
    ) %>%
    rename_if(
        .predicate = is.double,
        .funs = ~str_c("log10_", .)
    )

df_log %>%
    select_if(is.numeric) %>%
    ggpairs(progress = FALSE)
```

Now let's run a PCA to identify any low-dimensional structure.

```{r log_pca, echo=FALSE, message=FALSE}
do_pca <- function(df, ...) {
    pca_vars <- quos(...)

    df %>%
    filter_at(
        .vars = vars(!!! pca_vars),
        all_vars(!is.na(.))
    ) %>%
    tidy_pca(!!! pca_vars)
}

df_pca_log <-
    df_log %>%
    do_pca(
        log10_approach_speed,
        log10_wingspan,
        log10_length,
        log10_tail_height,
        log10_mtow
    )
```

```{r pca_report}
df_pca_log %>% pull(pc_frac)
```

About `87%` of the variance is explained by two directions; this is something we
can plot.

```{r fcn_plot_pca, echo=FALSE, message=FALSE}
do_pca_plot <- function(df, col_var) {
    col_var <- enquo(col_var)

    scale <- 4.0

    df_components <-
        df %>%
        pull(pca) %>%
        .[[1]] %>%
        .$rotation %>%
       as_tibble(rownames = "val")

    df %>%
        unnest(data_aug) %>%
        rename_all(~ str_replace(., patter = "\\.fitted", replacement = "")) %>%
        ggplot(aes(PC1, PC2, color = as.factor(!! col_var))) +
        geom_point() +
        geom_segment(
            data = df_components,
            aes(x = 0, y = 0, xend = PC1 * scale, yend = PC2 * scale),
            arrow = arrow(),
            color = "red"
        ) +
        geom_text_repel(
            data = df_components,
            aes(PC1 * scale, PC2 * scale, label = val),
            color = "red",
            box.padding = 1.0
        ) +
        theme(legend.position = "bottom")
}
```

```{r pca_plot}
df_pca_log %>%
    do_pca_plot(aac)
```

[Approach
speed](https://www.skybrary.aero/index.php/Approach_Speed_Categorisation) is
legally stratified into classes A-E; a fact that is visible in the figure above.
The first principal component is strongly dependent on the aircraft geometry,
while the second is quite aligned with the approach speed.

```{r approach_vs_engines}
df_faa %>%
    ggplot(aes(num_engines, approach_speed, group = num_engines)) +
    geom_boxplot()
```

As we might expect, the approach speed does vary with the number of engines.

```{r pca_2}
df_pca_2 <-
    df_log %>%
    do_pca(
        log10_wingspan,
        log10_length,
        log10_cmg,
        log10_mtow
    )

df_pca_2 %>% pull(pc_frac)

df_pca_2 %>%
    do_pca_plot(num_engines)
```
