---
title: "Dimension Reduction Examples"
author: "ZDR"
date: 2018-09-28
output:
  github_document:
    toc: true
urlcolor: blue
---

Examples of dimension reduction techniques applied to pipe flow problem.

## Setup
<!-- -------------------------------------------------- -->

```{r setup, echo=FALSE}
library(tidyverse)
library(piMat)
library(ggrepel)

## Script parameters
file_as_data0 <- "./as_data0.csv"
file_as_data1 <- "./as_data1.csv"
file_as_data2 <- "./as_data2.csv"

file_sobol_data2 <- "./sobol_data2.csv"

sel_idx <- 2

## Metadata
df_dim <-
  tribble(
    ##    rho,   U,   d,  mu, epsilon
    ~dim, ~Q1, ~Q2, ~Q3, ~Q4, ~Q5,
     "M",   1,   0,   0,   1,   0,
     "L",  -3,   1,   1,  -1,   1,
     "T",   0,  -1,   0,  -1,   0
  )

df_ref_basis <-
  tribble(
    ~out_name, ~Re, ~R,
         "Q1",   1,  0,
         "Q2",   1,  0,
         "Q3",   1, -1,
         "Q4",  -1,  0,
         "Q5",   0,  1
  )

## Common theme settings
theme_common <- function() {
  theme_minimal() %+replace%
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(4, 4, 4, 4), size = 16),
    axis.title.y = element_text(margin = margin(4, 4, 4, 4), size = 16, angle = 90),

    aspect.ratio = 1 / 1,

    plot.margin   = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),
    panel.spacing = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),

    legend.text = element_text(size = 10),

    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_line(color = "grey90")
  )
}
```

Load the data.

```{r load}
df_sobol_data <-
  read_csv(file_sobol_data2) %>%
  select(-X1)

df_as_data <-
  bind_rows(
    read_csv(file_as_data0) %>%
      select(-X1),
    read_csv(file_as_data1) %>%
      select(-X1),
    read_csv(file_as_data2) %>%
      select(-X1)
  )
```

## Sobol' indices
<!-- -------------------------------------------------- -->

Let's inspect the pre-computed Sobol' indices for the pipe flow problem.

```{r sobol}
df_sobol_data %>%
  filter(min_rank(-N) == 1) %>%
  select(-N) %>%
  mutate_all(~ round(., digits = 2) * 100) %>%
  knitr::kable(format = "latex")
```

These results indicate that the fifth input (roughness lengthscale) contributes
`95%` of the variance to the output (friction factor). In practice, we could fix
the first through fourth variables, and lose roughly `6%` of the variance in the
output. Depending on our application, this could be a negligible loss -- and a
huge win in terms of dimension reduction.

## Active subspace cartoon
<!-- -------------------------------------------------- -->

Build up a cartoon depiction of the active subspace.

```{r as_cartoon}
## Contours along
tibble(
  x = c(0, 0),
  y = c(0, 0),
  xend = c(0.5,  0.5),
  yend = c(0.5, -0.5),
  dir = c("Active", "Inactive")
) %>%
  ggplot() +
  geom_abline(slope = -1, intercept = 0) +
  geom_abline(slope = -1, intercept = -1) +
  geom_abline(slope = -1, intercept = -1.5) +
  geom_abline(slope = -1, intercept = +1) +
  geom_abline(slope = -1, intercept = +1.5) +
  geom_point(aes(x = 0, y = 0), color = "white") +
  coord_cartesian(xlim = c(-1, +1), ylim = c(-1, +1)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
  ) +
  labs(
    x = "x1",
    y = "x2"
  )

ggsave(filename = "../images/as_contour0.png")

## Dodged labels
tibble(
  x = c(0, 0),
  y = c(0, 0),
  xend = c(0.5,  0.5),
  yend = c(0.5, -0.5),
  dir = c("Active", "Inactive")
) %>%
  ggplot() +
  geom_abline(slope = -1, intercept = 0) +
  geom_abline(slope = -1, intercept = -1) +
  geom_abline(slope = -1, intercept = -1.5) +
  geom_abline(slope = -1, intercept = +1) +
  geom_abline(slope = -1, intercept = +1.5) +
  geom_segment(
    aes(x, y, xend = xend, yend = yend, color = dir),
    arrow = arrow()
  ) +
  geom_text(aes(xend, yend, label = dir), nudge_x = 0.1, nudge_y = 0.1) +
  coord_cartesian(xlim = c(-1, +1), ylim = c(-1, +1)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
  ) +
  labs(
    x = "x1",
    y = "x2"
  )

ggsave(filename = "../images/as_contour1.png")
```

## Vanilla active subspaces
<!-- -------------------------------------------------- -->

Estimate the active subspace.

```{r as_nat}
grad_mat_nat <-
  df_as_data %>%
  filter(bnd_idx == sel_idx) %>%
  select(G1:G5) %>%
  as.matrix()
as_mat_nat <-
  t(grad_mat_nat) %*% grad_mat_nat
as_svd_nat <-
  svd(as_mat_nat)

weights_leading_nat <-
  as_svd_nat$u[ , 1]
names(weights_leading_nat) <- str_c("Q", seq(1,5))
```

Assess the (relative) eigenvalue gaps.

```{r as_nat_gap}
(-diff(as_svd_nat$d) / sum(as_svd_nat$d)) %>%
  tibble(
    `Gap Fraction` = .,
    `Dimension` = seq(1, 4)
  ) %>%
  knitr::kable(format = "latex")
```

The spectrum suggests a 1-dimensional active subspace. Let's plot a summary.

```{r summary_nat}
df_as_data %>%
  filter(bnd_idx == sel_idx) %>%
  mutate(xi = named_inner(weights_leading_nat, df_as_data)) %>%
  ggplot(aes(xi, F)) +
  geom_point() +
  theme_common() +
  labs(
    x = "Active Variable 1",
    y = "Response"
  )
```

## Log-trick active subspaces
<!-- -------------------------------------------------- -->

Estimate the active subspace, but use dimensional analysis to inform this
operation.

```{r as_pi}
grad_mat_pi <-
  df_as_data %>%
  filter(bnd_idx == sel_idx) %>%
  mutate(
    G1p = Q1 * G1,
    G2p = Q2 * G2,
    G3p = Q3 * G3,
    G4p = Q4 * G4,
    G5p = Q5 * G5
  ) %>%
  select(G1p:G5p) %>%
  as.matrix()
as_mat_pi <-
  t(grad_mat_pi) %*% grad_mat_pi
as_svd_pi <-
  svd(as_mat_pi)

weights_leading_pi <-
  as_svd_pi$u[ , 1]
names(weights_leading_pi) <- str_c("Q", seq(1,5))
```

Assess the (relative) eigenvalue gaps.

```{r as_pi_gap}
-diff(as_svd_pi$d) / as_svd_pi$d[1]
```

The spectrum suggests a 1-dimensional active subspace. Let's plot a summary.

```{r summary_pi}
df_as_data %>%
  filter(bnd_idx == sel_idx) %>%
  mutate(
    xi = named_inner(
      weights_leading_pi,
      df_as_data %>%
      mutate_at(vars(Q1:Q5), log)
    )
  ) %>%
  ggplot(aes(xi, F)) +
  geom_point() +
  theme_common() +
  labs(
    x = "log(pi)",
    y = "Response"
  )
```

Re-express leading AS vector in terms of known dimensionless numbers.

```{r re_express}
df_given <-
  bind_rows(weights_leading_pi) %>%
  mutate(out_name = "pi1")

convert_space(transpose(df_given), df_ref_basis)
```
