---
title: "autos Live"
author: "The Wellesley Math Department"
date: 2018-09-06
output:
  github_document:
  toc: true
---
  
The following is a collaborative analysis of the [UCI](https://archive.ics.uci.edu/ml/datasets/automobile) `autos` 
dataset -- you'll need to download it to run this Rmarkdown file. The data are automobile characteristics and
(actuarial) risk assments for imported vehicles in 1985.

Also see the `backup_autos.Rmd` for a 'pre-baked' look at the same data.

```{r setup, echo=FALSE, warning=FALSE}
library(tidyverse)
library(tidyPca)   # Provides tidy_pca()
library(ggrepel)

## Data filename 
file_name <- "~/Sync/data/autos/imports-85.data" # Change to data location on your machine
## Variable names from the UCI site
var_names <- 
  c(
    "symboling",          # Actuarial risk adjustment: -3 Safest, +3 Riskiest
    "normalized_losses",  # Average insurance loss, averaged over body_style     
    "make",               # Manufacturer
    "fuel_type",          # Gas or diesel 
    "aspiration",         # Standard or turbo      
    "num_of_doors",       #      
    "body_style",         # hardtop, wagon, sedan, hatchback, convertible      
    "drive_wheels",       # Front, rear, all      
    "engine_location",    # Front, rear      
    "wheel_base",         # Distance between front & rear wheels      
    "length",             #      
    "width",              #      
    "height",             #      
    "curb_weight",        # Weight w/o cargo      
    "engine_type",        # Engine layout      
    "num_of_cylinders",   #      
    "engine_size",        # Total engine displacement      
    "fuel_system",        # Fuel feed system      
    "bore",               # Cylinder diameter      
    "stroke",             # Cylinder displacement distance      
    "compression_ratio",  # Swept volume compression ratio      
    "horsepower",         # Engine power      
    "peak_rpm",           # Peak rotations per minute (redline?)
    "city_mpg",           # City fuel efficiency      
    "highway_mpg",        # Highway fuel efficiency      
    "price"               # MSRP, presumably
  )

## Helper functions
## --------------------------------------------------
## Do a two-PC plot, colored by an additional variable
do_pca_plot <- function(df, col_var, components = TRUE, scale = 2.0) {
    col_var <- enquo(col_var)

    df_components <-
        df %>%
        pull(pca) %>%
        .[[1]] %>%
        .$rotation %>%
       as_tibble(rownames = "val")

    ## Start the plot
    fig <-
        df %>%
        unnest(data_aug) %>%
        rename_all(~ str_replace(., patter = "\\.fitted", replacement = "")) %>%
        ggplot()

    ## Optional color
    if (rlang::quo_is_missing(col_var)) {
        fig <-
            fig +
            geom_point(aes(PC1, PC2))

    } else {
        fig <-
            fig +
            geom_point(aes(PC1, PC2, color = as.factor(!! col_var)))

    }

    ## Optional component plot
    if (components) {
        fig <-
            fig +
            geom_segment(
                data = df_components,
                aes(x = 0, y = 0, xend = PC1 * scale, yend = PC2 * scale),
                ## arrow = arrow(),
                color = "red"
            )+
            geom_text_repel(
                data = df_components,
                aes(PC1 * scale, PC2 * scale, label = val),
                color = "red",
                box.padding = 1.0,
                segment.size = 0.1
            )
    }

    ## Finish and return the plot
    fig +
        theme(legend.position = "bottom") +
        labs(
            x = "Principal Component Score 1",
            y = "Principal Component Score 2"
        )
}

## Parse character -> integer
parse_numchar <- function(x) {
  case_when(
        x == "two" ~ 2L,
        x == "three" ~ 3L,
        x == "four" ~ 4L,
        x == "five" ~ 5L,
        x == "six" ~ 6L,
        x == "seven" ~ 7L,
        x == "eight" ~ 8L,
        x == "nine" ~ 9L,
        x == "ten" ~ 10L,
        x == "eleven" ~ 11L,
        x == "twelve" ~ 12L,
        TRUE ~ NA_integer_
      )
}
```

## Wrangling
<!-- ------------------------------------------------- -->

The data are essentially in CSV format, though we'll need to do some wrangling.

```{r load_n_wrangle}
df_raw <- 
  read_csv(
    file_name, 
    col_names = var_names,
    col_types = cols(
      .default = col_number(),
      symboling = col_integer(),
      normalized_losses = col_integer(),
      make = col_character(),
      fuel_type = col_character(),
      aspiration = col_character(),
      num_of_doors = col_character(),     # Needs manual fix!
      body_style = col_character(),
      drive_wheels = col_character(),
      engine_location = col_character(),
      engine_type = col_character(),
      num_of_cylinders = col_character(), # Also needs manual fix!
      engine_size = col_integer(),
      fuel_system = col_character()
    )
  )
```

The string-coded numbers are obnoxious; let's map those to integers.

```{r}
df_raw %>% pull(num_of_doors) %>% unique()
df_raw %>% pull(num_of_cylinders) %>% unique()
```


```{r}
df_autos <- 
  df_raw %>% 
  mutate(
    num_of_doors = parse_numchar(num_of_doors),
    num_of_cylinders = parse_numchar(num_of_cylinders)
  )
```


## Analysis
<!-- ------------------------------------------------- -->

```{r}
df_autos %>% 
  glimpse()
```

## Questions
<!-- ------------------------------------------------- -->

__q1__ What is the relationship between the geometric variables `length`, `width`, `height`, and `wheel_base`?

```{r}
pca_geo <- 
  df_autos %>% 
  tidy_pca(length, width, height, wheel_base)

pca_geo %>% 
  pull(pc_frac)
```

At two dimensions, we capture ~`79`% of the variability in the data -- this is usable low-dimensional structure.

```{r}
pca_geo %>%
  do_pca_plot(col_var = body_style)
```

Interpreting the PCA directions, we see that `height`, `wheel_base`, `length`, and `width` are all positively
related with the first direction -- this first direction seems to correspond to vehicle size.

The second direction is positively related with `height`, and negatively related with `width` -- this is a sort
of aspect ratio.

Plotting body type on the PCA scores, we see that sedans tend to be scattered along the 1st direction,
will little variation on the second. It seems these sedans come in different sizes, but do not vary
much in shape (aspect ratio).

__q2__ What is the relationship between the engine variables `engine_size`, `bore`, `stroke`, `compression_ratio`, and `num_of_cylinders`?

We suspect all of these variables are positively correlated, as the all pertain to the size of the engine. We can start by analying
a subset of them

```{r}
pca_engine <- 
  df_autos %>% 
  filter_at(vars(engine_size, bore, stroke), all_vars(!is.na(.))) %>% 
  tidy_pca(engine_size, bore, stroke)

pca_engine %>% 
  pull(pc_frac)
```

We can capture most of the variability with two out of three dimensions.

```{r}
pca_engine %>% 
  do_pca_plot()
```

It's a bit inconvenient, but the engine size is negatively correlated with the first PCA direction -- the first direction depends primarily
on engine size and bore. The second direction depends primarily on stroke.

__q3__ What about dimensionality made these questions hard?

__q4__ What else is going on in these data?
